The effects of probabilistic context inference on motor adaptation

### 2022-06-27 adaptation/article/article.tex
Update on Overleaf.
Final draft for re-submission

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
%% \documentclass{article}
\usepackage[font=large]{caption}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{todonotes}
\usepackage{authblk}
% \usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{multirow}
\usepackage{tabularx}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when sudden or gradual changes to the dynamics of the environment are introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Humans are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. Recently, the COIN computational model for motor adaptation and context inference was introduced, which contains explicit components of context inference and Bayesian motor adaptation. This model was used to show the effects of context inference on learning rates across different experiments. We expanded on that work by using the COIN model to show that the effects of context inference on motor adaptation and control go even further than previously shown. Here, we used the COIN model to simulate classical motor adaptation experiments from previous works and showed that context inference, and how it is affected by the presence and reliability of feedback, effect a host of behavioral phenomena that had so far required multiple hypothesized mechanisms, lacking a unified explanation. Concretely, we show that the reliability of direct contextual information, as well as noisy sensory feedback, typical of many experiments, effect measurable changes in switching-task behavior, as well as in action selection, that stem directly from probabilistic context inference.

@@

\subsubsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Spontaneous recovery, during which participants' behavior is consistent with a previously-encountered context (e.g. in an O-A-E experiment, consistent with A); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline, which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of the spontaneous recovery, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the adaptation vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.

@@

Context inference also explains the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). The start and speed of the drop depend on the run (i.e. the participant), reproducing the different lags observed in experimental data.

@@

The same rationale explains the results by \cite{Herzfeld_memory_2014}: motor learning, which in the COIN model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1). Additionally, as was shown by \cite{Heald_Contextual_2021}, context inference explains the modulation of learning rate by the volatility of the environment observed by \cite{Herzfeld_memory_2014}.

@@

\subsubsection{Error-clamp as a known context}
The inclusion of reliable sensory contextual cues (e.g. lights whose color uniquely identify a context) makes switching immediate, as in the experiments by \cite{Kim_Neural_2015}. We expect that the same effect would be observed in error-clamp trials. If the E block is learned by participants during training, it might still be difficult for them to infer that an E block has started, which would create delays similar to those in \fref{fig:vaswani-2013}. However, the model predicts that if a visual cue is introduced that identifies the E block, participants would immediately switch to their baseline behavior, no longer displaying spontaneous recovery, lag, nor the slow return to baseline. This immediate switch in the presence of contextual cues would persist even if endpoint feedback is manipulated as \cite{Vaswani_Decay_2013} did.

Note that the original COIN specification includes a component to learn new contexts. However, this component works exclusively by creating new contexts in which the forward models take the same form but have different parameter values. It is currently unclear whether the model can be adapted to create contexts in an online fashion that operate in an essentially different manner, as is the case of error-clamp trials, in which participants' responses do not affect the outcome and motor commands are issued based on criteria not directly related to the goal of the task (e.g. energy minimization or comfort maximization).

\subsubsection{Interference effects}
As discussed in the Resuls section, the effect observed by \cite{Davidson_Scaling_2004} is explained by the model as an effect of slow context inference, instead of being a direct interference at the level of learning. As shown in \fref{fig:davidson-2004}B-C, the context inference account predicts that this effect would disappear if all contexts were significantly different from baseline, such that the baseline context never explains the observations. Removing the baseline context from a participant's context inference might be experimentally unfeasible, but other possibilities include making all adaptations bigger (e.g. bigger angles, stronger forces), and including contextual cues that rule out the baseline context. In the opposite direction, the model predicts that if all adaptations are smaller (i.e. closer to baseline), the differences between the two groups would increase.

>>>>

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
%% \documentclass{article}
\usepackage[font=large]{caption}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{todonotes}
\usepackage{authblk}
% \usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{lineno}
\linenumbers

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when sudden or gradual changes to the dynamics of the environment are introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Humans are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. Recently, the COIN computational model for motor adaptation and context inference was introduced, which contains components for context inference and Bayesian motor adaptation. This model was used to show the effects of context inference on learning rates across different experiments. We expanded on that work by using the COIN model to show that the effects of context inference on motor adaptation and control go even further than previously shown. Here, we used the COIN model to simulate classical motor adaptation experiments from previous works and showed that context inference, and how it is affected by the presence and reliability of feedback, effect a host of behavioral phenomena that had so far required multiple hypothesized mechanisms, lacking a unified explanation. Concretely, we show that the reliability of direct contextual information, as well as noisy sensory feedback, typical of many experiments, effect measurable changes in switching-task behavior, as well as in action selection, that stem directly from probabilistic context inference.

@@

\subsubsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Participants' behavior is consistent with a previously-encountered context (called spontaneous recover in O-A-B-E experiments, where behavior is consistent with A); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline, which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of adaptation during the first phase, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the response vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.

@@

Context inference also explains the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. Furthermore, variability in the lag between and within groups is explained by perceptual noise, which directly affects context inference. To show these effects, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). The start and speed of the drop depend on the run (i.e. the participant), reproducing the different lags observed in experimental data.

@@

The same rationale explains the results by \cite{Herzfeld_memory_2014}, as was shown by \cite{Heald_Contextual_2021}: motor learning, which in the COIN model is modulated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1). Additionally, as was shown by \cite{Heald_Contextual_2021}, context inference explains the modulation of learning rate by the volatility of the environment observed by \cite{Herzfeld_memory_2014}.

@@

For the model predictions discussed below, it is important to keep in mind that different contextual cues are not equally effective at separating motor responses during learning and switching \cite{Howard_Gone_2012,Howard_ContextDependent_2010,Imamizu_Explicit_2007}. Because of this, the model predictions hinge on selecting the adequate type of contextual information that maximally helps the participants select the appropriate motor response.

\subsubsection{Error-clamp as a known context}
The inclusion of reliable sensory contextual cues (e.g. lights whose color uniquely identify a context) makes switching immediate, as in the experiments by \cite{Kim_Neural_2015}. We expect that the same effect would be observed in error-clamp trials. If the E block is learned by participants during training, it might still be difficult for them to infer that an E block has started, which would create delays similar to those in \fref{fig:vaswani-2013}. However, the model predicts that if a visual cue is introduced that identifies the E block, participants would immediately switch to their baseline behavior, no longer displaying an adapted response, lag, nor the slow return to baseline. This immediate switch in the presence of contextual cues would persist even if endpoint feedback is manipulated as \cite{Vaswani_Decay_2013} did.

Note that the original COIN specification includes a component to learn new contexts. However, this component works exclusively by creating new contexts in which the forward models take the same form but have different parameter values. New mechanisms would be needed to allow the COIN model to create contexts in an online fashion that operate in an essentially different manner, as is the case of error-clamp trials, in which participants' responses do not affect the outcome and motor commands are issued based on criteria not directly related to the goal of the task (e.g. energy minimization or comfort maximization).

\subsubsection{Interference effects during context switching}
As discussed in the Resuls section, the effect observed by \cite{Davidson_Scaling_2004} is explained by the model as an effect of slow context inference, instead of being a direct interference at the level of learning. As shown in \fref{fig:davidson-2004}B-C, the context inference account predicts that this effect would disappear if all contexts were significantly different from baseline, such that the baseline context never explains the observations. Removing the baseline context from a participant's context inference might be experimentally unfeasible, but other possibilities include making all adaptations bigger (e.g. bigger angles, stronger forces), and including contextual cues that rule out the baseline context. In the opposite direction, the model predicts that if all adaptations are smaller (i.e. closer to baseline), the differences between the two groups would increase.
